Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2023-12-01 manuscript/manuscript.Rmd
Switching to msystems fig width of 6.875 inches

library(egg)
nolegend_plot <- ggarrange(nmds_sla_gp+theme(legend.position = "none"), 
                           med_metab_sla_gp, ncol = 1, heights = c(0.7, 0.3), 
                           draw = FALSE)
library(cowplot)
nmds_and_med_metab <- plot_grid(nolegend_plot, get_legend(nmds_sla_gp), 
                                nrow=1, rel_widths = c(0.9, 0.1))
ggsave("nmds_and_med_metab.png", plot = nmds_and_med_metab, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300,
       bg="white")
```

@@

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth, tripl)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot(aes(label=tripl)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=4, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=4, pch=21, color="black") +
  scale_fill_manual(
    values = c("grey90", "grey80", "grey60", "grey40", "grey20"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.2, 0, 0.2, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal"
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.5, height = 4.5, units = "in", dpi = 300)
```

@@

```{r kclust volcano plot combine}
kclust_volcano_gp <- egg::ggarrange(
  kclust_gp, volcanoplot_gp + theme(legend.position = "none"), 
  ncol=1, heights = c(2, 1.5)) %>%
  cowplot::plot_grid(
    cowplot::get_legend(volcanoplot_gp), rel_widths = c(0.85, 0.15)
  )
ggsave("kclust_volcano_gp.png", plot = kclust_volcano_gp, device = "png", 
       path = "figures", width = 6.5, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

ggsave("fk_nmdsplot.png", plot = fk_nmdsplot, device = "png",
       path = "figures", width = 6.5, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

lm_gp <- ggplot(pcdata, aes(x=sla_corr, y=value)) +
  geom_smooth(method="lm", formula=y~x, color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  geom_label(data = pcvars, aes(label=perc_label),
             hjust=1, vjust=1, x=Inf, y=Inf, label.r = unit(0, "in")) +
  facet_grid(depth~name, switch = "y") +
  scale_y_continuous(position = "right") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"), guide = NULL
  ) +
  labs(x="Corrected SLA, in cm", y="Principal component value",
       color="Corrected\nSLA (cm)") +
  theme_bw()
deep_pc1_barplot_gp <- pcdata %>%
  left_join(filled_file_metadata %>% distinct(filename, station, tripl)) %>%
  mutate(station=factor(station, levels=as.character(4:14),
                        labels=paste0("St.", 4:14))) %>%
  arrange(station) %>%
  filter(name=="PC1" & str_detect(depth, "Deep")) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=tripl, y=value)) +
  facet_wrap(~station, nrow = 1, scales="free_x") +
  scale_y_continuous(position = "right") +
  labs(x="Triplicate", y="PC1", title = paste(
    "    Variance among deep (175 meter) samples in PC1"
  ))
pca_sla_cor_gp <- cowplot::plot_grid(lm_gp, deep_pc1_barplot_gp, 
                                     rel_heights = c(0.6, 0.4), 
                                     ncol = 1, labels = "AUTO")
ggsave("pca_sla_cor_gp.png", pca_sla_cor_gp, device = "png", path = "supplement",
       width = 6.5, height = 6.5, units = "in", dpi = 300)

@@

```{r supp figure fraction of PC in metabolites}
pc_comp_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

pc_frac_metab_gp <- ggplot(pc_comp_df, aes(x=station, y=frac_in_metab, color=sla_corr)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.9, linewidth=0.8) +
  geom_point(color="black", size=2) +
  scale_y_continuous(labels = scales::label_percent()) +
  facet_grid(~depth, scales="free") +
  scale_color_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill")
  ) +
  guides(color=guide_colorbar(direction = "horizontal", title.position = "top", 
                              title.hjust = 0.5)) +
  labs(color="Sea level\nanomaly (cm)", x="Station",
       y="Fraction of PC represented by known metabolites") +
  theme_bw() +
  theme(legend.position = c(0.85, 0.8), legend.background = element_rect(color="grey90"))
ggsave("pc_frac_metab_gp.png", pc_frac_metab_gp, device = "png", path = "supplement",
       width = 6.5, height = 3.5, units = "in", dpi = 300)
```

![](supplement/pc_frac_metab_gp.png)

@@

ggsave("fk_nmds_and_med_metab.png", plot = fk_nmds_and_med_metab, device = "png",
       path = "supplement", width = 6.5, height = 5.5, units = "in", dpi = 300,
       bg="white")
```

@@

```{r possible figure stacked barplot ranked space}
top10_stacked_area_station_gp <- map(c("Surface", "DCM", "Deep"), function(depth_i){
  reg_coefs %>%
    filter(str_detect(term, "sla_corr")) %>%
    mutate(p.value=p.adjust(p.value, method="fdr")) %>%
    filter(depth==depth_i) %>%
    arrange(p.value) %>%
    group_by(estimate>0) %>%
    slice(1:5) %>%
    ungroup() %>%
    left_join(untarg_feats %>% select(feature, compound_name, mzmed, rtmed)) %>%
    mutate(feature=ifelse(!is.na(compound_name), compound_name, 
                          paste0(round(mzmed, 3), "_", round(rtmed, 1)))) %>%
    mutate(feature=fct_inorder(feature)) %>%
    unnest(data) %>%
    select(feature, depth, norm_area, filename, station) %>%
    group_by(feature) %>%
    # mutate(norm_area=rank(norm_area)) %>%
    mutate(norm_area=norm_area/sum(norm_area)) %>%
    mutate(station=factor(station, levels=as.character(4:14))) %>%
    ggplot() +
    geom_col(aes(x=station, y=norm_area, fill=feature), position = "fill") +
    scale_y_continuous(expand = c(0,0)) +
    labs(x="Station", y="Relative peak area", fill="Mass feature") +
    theme_bw()
}) %>% gridExtra::arrangeGrob(grobs = ., ncol = 1)
ggsave("top10_stacked_area_station_gp.png", plot = top10_stacked_area_station_gp, 
       device = "png", path = "supplement", width = 6.5, height = 9, 
       units = "in", dpi = 300)
```

@@

ggsave("katherine_comp_mostcycenriched_gp.png", plot = katherine_comp_mostcycenriched_gp, 
       device = "png", path = "supplement", width = 6.5, height = 5, units = "in", dpi = 300)
```

@@

```{r enrichment plot DCM comparison, eval=FALSE}
MC_stacked_mostdiff_gp <- pairwise_tests %>%
  arrange(p.value) %>%
  group_by(statistic>110) %>%
  slice(1:5) %>%
  ungroup() %>%
  # filter(p.value<0.00005) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Dimethylsulfoniopropionate", "DMSP", compound_name)) %>%
  arrange(statistic) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  select(feature, compound_name) %>%
  left_join(untarg_peaks %>% select(filename, norm_area, feature), by="feature") %>%
  left_join(filled_file_metadata %>% distinct(filename, station, abs_depth), by="filename") %>%
  filter(station%in%c("L1", "L2")) %>%
  arrange(station) %>%
  mutate(station=ifelse(station=="L1", "Cyclone", "Anticyclone")) %>%
  mutate(station=fct_inorder(station)) %>%
  arrange(abs_depth) %>%
  mutate(filename=fct_inorder(filename)) %>%
  group_by(feature) %>%
  # mutate(norm_area=rank(norm_area)) %>%
  mutate(norm_area=norm_area/sum(norm_area)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(x=filename, y=norm_area, fill=compound_name, group=filename), 
           position = "fill", color="grey50", width=1) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=1, ymax=1.1), 
            fill=c("#8277b6", "#b05d56"), color="black",
            data = data.frame(station=fct_inorder(c("Cyclone", "Anticyclone")))) +
  geom_text(aes(x=8, y=1.05, label=station), hjust=0.5,
            data = data.frame(station=fct_inorder(c("Cyclone", "Anticyclone")))) +
  facet_grid(~station, scales="free_x") +
  scale_fill_manual(values = splividis(10, begin = 0.1, end = 0.95)) +
  guides(fill=guide_legend(reverse = TRUE)) +
  scale_y_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background = element_blank(), strip.text = element_blank()) +
  labs(x=NULL, y="Relative peak area", fill="Mass feature")
ggsave("MC_stacked_mostdiff_gp.png", plot = MC_stacked_mostdiff_gp, device = "png", 
       path = "supplement", width = 6.5, height = 3, units = "in", dpi = 300)
```

>>>>

library(egg)
nolegend_plot <- ggarrange(nmds_sla_gp+theme(legend.position = "none"), 
                           med_metab_sla_gp, ncol = 1, heights = c(0.7, 0.3), 
                           draw = FALSE)
library(cowplot)
nmds_and_med_metab <- plot_grid(nolegend_plot, get_legend(nmds_sla_gp), 
                                nrow=1, rel_widths = c(0.9, 0.1))
ggsave("nmds_and_med_metab.png", plot = nmds_and_med_metab, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300,
       bg="white")
```

@@

```{r figure MC NMDS plot}
MC_nmds_gp <- multivar_analysis_MC$nmds_output %>%
  pluck(1, "points") %>% 
  as.data.frame() %>%
  rownames_to_column("filename") %>%
  left_join(filled_file_metadata %>% distinct(filename, sla_class, depth, tripl)) %>%
  mutate(depth=factor(depth, levels=c("DCMless20m", "DCMless10m", "DCM", 
                                      "DCMplus10m", "DCMplus20m"),
                      labels=c("-20m", "-10m", "DCM", "+10m", "+20m"))) %>%
  ggplot(aes(label=tripl)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point(aes(x=MDS1, y=MDS2, color=sla_class), size=4, pch=19) +
  scale_color_manual(
    values=c(scales::muted("blue"), scales::muted("red")),
    breaks=c("Cyclone", "Anticyclone"),
  ) +
  annotate("label", x=-200, y=100, label="Cyclone", color=scales::muted("blue"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=100, y=120, label="Anticyclone", color=scales::muted("red"),
           label.r = unit(0, "in"), size=6) +
  annotate("label", x=-Inf, y=Inf, label=multivar_analysis_MC$nmds_stress,
           hjust=0, vjust=1, label.r = unit(0, "in")) +
  geom_point(aes(x=MDS1, y=MDS2, fill=depth, alpha=depth), size=4, pch=21, color="black") +
  scale_fill_manual(
    values = c("grey90", "grey80", "grey60", "grey40", "grey20"),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  scale_alpha_manual(
    values = c(0.4, 0.2, 0, 0.2, 0.4),
    breaks = c("-20m", "-10m", "DCM", "+10m", "+20m")
  ) +
  coord_fixed() +
  guides(alpha="none", color="none", fill=guide_legend(
    title = "Depth", title.position = "top", title.hjust = 0.5, direction = "horizontal"
  )) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave("MC_nmds_gp.png", plot = MC_nmds_gp, device = "png", 
       path = "figures", width = 6.875, height = 4.5, units = "in", dpi = 300)
```

@@

```{r kclust volcano plot combine}
kclust_volcano_gp <- egg::ggarrange(
  kclust_gp, volcanoplot_gp + theme(legend.position = "none"), 
  ncol=1, heights = c(2, 1.5)) %>%
  cowplot::plot_grid(
    cowplot::get_legend(volcanoplot_gp), rel_widths = c(0.85, 0.15)
  )
ggsave("kclust_volcano_gp.png", plot = kclust_volcano_gp, device = "png", 
       path = "figures", width = 6.875, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

ggsave("fk_nmdsplot.png", plot = fk_nmdsplot, device = "png",
       path = "figures", width = 6.875, height = 4, units = "in", dpi = 300,
       bg="white")
```

@@

lm_gp <- ggplot(pcdata, aes(x=sla_corr, y=value)) +
  geom_smooth(method="lm", formula=y~x, color="black") +
  geom_point(aes(fill=sla_corr), pch=21, color="black") +
  geom_label(data = pcvars, aes(label=perc_label),
             hjust=1, vjust=1, x=Inf, y=Inf, label.r = unit(0, "in")) +
  facet_grid(depth~name, switch = "y") +
  scale_y_continuous(position = "right") +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"),  mid = "grey90",
    aesthetics = c("color", "fill"), guide = NULL
  ) +
  labs(x="Corrected SLA, in cm", y="Principal component value",
       color="Corrected\nSLA (cm)") +
  theme_bw()
deep_pc1_barplot_gp <- pcdata %>%
  left_join(filled_file_metadata %>% distinct(filename, station, tripl)) %>%
  mutate(station=factor(station, levels=as.character(4:14),
                        labels=paste0("St.", 4:14))) %>%
  arrange(station) %>%
  filter(name=="PC1" & str_detect(depth, "Deep")) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_col(aes(x=tripl, y=value)) +
  facet_wrap(~station, nrow = 1, scales="free_x") +
  scale_y_continuous(position = "right") +
  labs(x="Triplicate", y="PC1", title = paste(
    "    Variance among deep (175 meter) samples in PC1"
  ))
pca_sla_cor_gp <- cowplot::plot_grid(lm_gp, deep_pc1_barplot_gp, 
                                     rel_heights = c(0.6, 0.4), 
                                     ncol = 1, labels = "AUTO")
ggsave("pca_sla_cor_gp.png", pca_sla_cor_gp, device = "png", path = "supplement",
       width = 6.875, height = 6.5, units = "in", dpi = 300)

@@

@@

ggsave("fk_nmds_and_med_metab.png", plot = fk_nmds_and_med_metab, device = "png",
       path = "supplement", width = 6.875, height = 5.5, units = "in", dpi = 300,
       bg="white")
```

@@

```{r possible figure stacked barplot ranked space}
top10_stacked_area_station_gp <- map(c("Surface", "DCM", "Deep"), function(depth_i){
  reg_coefs %>%
    filter(str_detect(term, "sla_corr")) %>%
    mutate(p.value=p.adjust(p.value, method="fdr")) %>%
    filter(depth==depth_i) %>%
    arrange(p.value) %>%
    group_by(estimate>0) %>%
    slice(1:5) %>%
    ungroup() %>%
    left_join(untarg_feats %>% select(feature, compound_name, mzmed, rtmed)) %>%
    mutate(feature=ifelse(!is.na(compound_name), compound_name, 
                          paste0(round(mzmed, 3), "_", round(rtmed, 1)))) %>%
    mutate(feature=fct_inorder(feature)) %>%
    unnest(data) %>%
    select(feature, depth, norm_area, filename, station) %>%
    group_by(feature) %>%
    # mutate(norm_area=rank(norm_area)) %>%
    mutate(norm_area=norm_area/sum(norm_area)) %>%
    mutate(station=factor(station, levels=as.character(4:14))) %>%
    ggplot() +
    geom_col(aes(x=station, y=norm_area, fill=feature), position = "fill") +
    scale_y_continuous(expand = c(0,0)) +
    labs(x="Station", y="Relative peak area", fill="Mass feature") +
    theme_bw()
}) %>% gridExtra::arrangeGrob(grobs = ., ncol = 1)
ggsave("top10_stacked_area_station_gp.png", plot = top10_stacked_area_station_gp, 
       device = "png", path = "supplement", width = 6.875, height = 9, 
       units = "in", dpi = 300)
```

@@

ggsave("katherine_comp_mostcycenriched_gp.png", plot = katherine_comp_mostcycenriched_gp, 
       device = "png", path = "supplement", width = 6.875, height = 5, units = "in", dpi = 300)
```

@@

```{r enrichment plot DCM comparison, eval=FALSE}
MC_stacked_mostdiff_gp <- pairwise_tests %>%
  arrange(p.value) %>%
  group_by(statistic>110) %>%
  slice(1:5) %>%
  ungroup() %>%
  # filter(p.value<0.00005) %>%
  left_join(untarg_feats %>% select(feature, compound_name), by = join_by(feature)) %>%
  mutate(compound_name=ifelse(is.na(compound_name), feature, compound_name)) %>%
  mutate(compound_name=ifelse(compound_name=="Dimethylsulfoniopropionate", "DMSP", compound_name)) %>%
  arrange(statistic) %>%
  mutate(compound_name=fct_inorder(compound_name)) %>%
  select(feature, compound_name) %>%
  left_join(untarg_peaks %>% select(filename, norm_area, feature), by="feature") %>%
  left_join(filled_file_metadata %>% distinct(filename, station, abs_depth), by="filename") %>%
  filter(station%in%c("L1", "L2")) %>%
  arrange(station) %>%
  mutate(station=ifelse(station=="L1", "Cyclone", "Anticyclone")) %>%
  mutate(station=fct_inorder(station)) %>%
  arrange(abs_depth) %>%
  mutate(filename=fct_inorder(filename)) %>%
  group_by(feature) %>%
  # mutate(norm_area=rank(norm_area)) %>%
  mutate(norm_area=norm_area/sum(norm_area)) %>%
  ungroup() %>%
  ggplot() +
  geom_col(aes(x=filename, y=norm_area, fill=compound_name, group=filename), 
           position = "fill", color="grey50", width=1) +
  geom_rect(aes(xmin=-Inf, xmax=Inf, ymin=1, ymax=1.1), 
            fill=c("#8277b6", "#b05d56"), color="black",
            data = data.frame(station=fct_inorder(c("Cyclone", "Anticyclone")))) +
  geom_text(aes(x=8, y=1.05, label=station), hjust=0.5,
            data = data.frame(station=fct_inorder(c("Cyclone", "Anticyclone")))) +
  facet_grid(~station, scales="free_x") +
  scale_fill_manual(values = splividis(10, begin = 0.1, end = 0.95)) +
  guides(fill=guide_legend(reverse = TRUE)) +
  scale_y_discrete(expand = c(0,0)) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.background = element_blank(), strip.text = element_blank()) +
  labs(x=NULL, y="Relative peak area", fill="Mass feature")
ggsave("MC_stacked_mostdiff_gp.png", plot = MC_stacked_mostdiff_gp, device = "png", 
       path = "supplement", width = 6.875, height = 3, units = "in", dpi = 300)
```
